### 2021-06-01 thesis/tex/sterileTheory.tex
update

% \input{commands}\begin{document}
\section{The Sterile State}\label{sec:anomalies}
In 1996, the LSND experiment reported an excess of $\ane$ events from an $\anm$ beam~\cite{lsnd}. The anomaly was consistent with a fourth
neutrino state with $\dm[41] > \SI{0.03}{\eV \squared}$. Nine years later, MiniBooNE not only reproduced the $\ane$ anomaly, but saw the excess in The
$\ne$ events too. Together with the so-called reactor and gallium analomalies, there were indications that both appearance and disappearance
anomalies might be remedied by a fourth mass state.

However, we know from the decay width of the $Z$ boson that it only can interact with three flavor species, so this fourth mass state can't be interacting weakly.
We now distinguish between the three original neutrino flavors ($e$, $\mu$, and $\tau$) and the new fourth flavor ($s$) by calling the former \emph{active} neutrinos
and the latter \emph{sterile}. The experiments listed above indicate that the mass-squared difference of the sterile neutrino is in the \si{\eV} scale, while the two others
are three and five magnitudes smaller. To remind us of this difference, we write $3+1$, since the sterile state proposed is a lot more massive than the active states. Models
with different hierachies might be written as $1+2+1$ or $1+3$, but many of them are ruled out by cosmological constraints and will not be considered here. %TODO:source

\subsection{The Hamiltonian}
The inclusion of the sterile neutrino in the Hamiltonian is straightforward. We extend the PMNS matrix to incorporate the new flavor and mass eigenstates: 
\begin{align}
    U_{4gen} =
    \begin{pmatrix}
    U_{e1} & U_{e2} & U_{e3} & U_{e4} \\
    U_{\mu1} & U_{\mu2} & U_{\mu3} & U_{\mu4} \\
    U_{\tau1} & U_{\tau2} & U_{\tau3} & U_{\tau4} \\
    U_{s1} & U_{s2} & U_{s3} & U_{s4}
    \end{pmatrix}\,,
\end{align}
where a common parametrization of the PMNS matrix is 
\begin{align}\label{eq:U4gen_param}
    U_{4gen} = R_{34}R_{24}R_{14}U_{3gen}\,.
\end{align}
The mass matrix extends analogously:
\begin{align}\label{eq:dm4gen}
    M^2_{4gen} =
    \begin{pmatrix}
        0 & 0 & 0 & 0\\
        0 & \dm_{21} & 0  & 0 \\
        0 & 0 & \dm_{31} & 0 \\
        0 & 0 & 0 & \dm_{41} \\
    \end{pmatrix}\,.
\end{align}

@@

\begin{align}\label{eq:Vsterile}
    V_{4gen} &=
    \begin{pmatrix}
        V_{CC} + V_{NC} & 0 & 0 & 0 \\
        0 &V_{NC} & 0 & 0 \\
        0 & 0 & V_{NC} & 0 \\
        0 & 0 & 0 & 0 
    \end{pmatrix} - V_{NC}\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 1 
    \end{pmatrix}
    = \begin{pmatrix}
        V_{CC} & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -V_{NC}
    \end{pmatrix} \nonumber \\
    &=\sqrt{2}G_F N_e\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -1/2
    \end{pmatrix}\,.
\end{align}
where we have assumed electrical neutrality in the last step, yielding $N_e = N_n$.

@@

So for a \si{\eV}-scale sterile neutrino and a non-zero $\theta_{24}$, we expect a \si{\TeV} $\anm$ disappearance.
The resonance dip is affected by the value of $\dm[41]$ and $\theta_{24}$ as shown in Fig.~\ref{fig:resonance_shift}.
We see how the value of $\dm[41]$ shifts the peak, while the mixing angle adjusts its strength. 
TODO: more? mention that Pmm=1 at TeV for 3+0

>>>>

% \input{commands}\begin{document}
\section{The Sterile State}\label{sec:anomalies}
In 1996, the LSND experiment reported an excess of $\ane$ events from an $\anm$ beam~\cite{lsnd}.  
Nine years later, MiniBooNE not only reproduced the $\ane$ anomaly, but saw the excess in The
$\ne$ events too. Together with the so-called reactor and gallium anomalies, these reports suggested 
that the three massive neutrino framework could be amended.
By introducing a fourth heavy mass state $\nu_4$, both appearance and disappeareance anomalies could be minimally accomodated.

However, we know from the decay width of the $Z$ boson that it only can interact with three flavor species, 
so this fourth mass state cannot be interacting weakly.
We now distinguish between the three original neutrino flavors ($e$, $\mu$, and $\tau$) and the new fourth 
flavor ($s$) by calling the former \emph{active} neutrinos
and the latter \emph{sterile}. The experiments listed above indicate that the mass-squared difference of the 
sterile neutrino is in the \si{\eV} scale, while the two others
are three and five magnitudes smaller. To remind us of this difference, we write $3+1$, since the sterile state 
proposed is a lot more massive than the active states. Models
with different hierachies might be written as $1+2+1$ or $1+3$, but many of them are ruled out by cosmological 
observations and will not be considered here. %TODO:source

In this chapter, we will use the matter-enhanced resonant oscillations assuming the 3+1 framework,
and probe it the using the $\nm$ disappearance signature.

\subsection{The Hamiltonian}
The inclusion of the sterile state in the Hamiltonian is straightforward. We extend the PMNS matrix to incorporate the new flavor and mass eigenstates: 
\begin{align}
    U_{4gen} =
    \begin{pmatrix}
    U_{e1} & U_{e2} & U_{e3} & U_{e4} \\
    U_{\mu1} & U_{\mu2} & U_{\mu3} & U_{\mu4} \\
    U_{\tau1} & U_{\tau2} & U_{\tau3} & U_{\tau4} \\
    U_{s1} & U_{s2} & U_{s3} & U_{s4}
    \end{pmatrix}\,,
\end{align}
where a common parametrization of this new mixing matrix is 
\begin{align}\label{eq:U4gen_param}
    U_{4gen} = R_{34}R_{24}R_{14}U_{3gen}\,.
\end{align}
The mass matrix extends analogously:
\begin{align}\label{eq:dm4gen}
    M^2_{4gen} =
    \begin{pmatrix}
        0 & 0 & 0 & 0\\
        0 & \dm_{21} & 0  & 0 \\
        0 & 0 & \dm_{31} & 0 \\
        0 & 0 & 0 & \dm_{41} \\
    \end{pmatrix}\,.
\end{align}

@@

\begin{align}\label{eq:Vsterile}
    V_{4gen} &=
    \begin{pmatrix}
        V_{CC} + V_{NC} & 0 & 0 & 0 \\
        0 &V_{NC} & 0 & 0 \\
        0 & 0 & V_{NC} & 0 \\
        0 & 0 & 0 & 0 
    \end{pmatrix} - V_{NC}\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &1 & 0 & 0 \\
        0 & 0 & 1 & 0 \\
        0 & 0 & 0 & 1 
    \end{pmatrix}
    = \sqrt{2}G_F\begin{pmatrix}
        N_e & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -N_n/2
    \end{pmatrix} \nonumber \\
    &=\sqrt{2}G_F N_e\begin{pmatrix}
        1 & 0 & 0 & 0 \\
        0 &0 & 0 & 0 \\
        0 & 0 & 0 & 0 \\
        0 & 0 & 0 & -1/2
    \end{pmatrix}\,.
\end{align}
where we have assumed electrical neutrality in the last step, yielding $N_e = N_n$.

@@

So for a \si{\eV}-scale sterile neutrino and a non-zero $\theta_{24}$, we expect a \si{\TeV} $\anm$ disappearance.
The resonance dip is affected by the value of $\dm[41]$ and $\theta_{24}$ as shown in Fig.~\ref{fig:resonance_shift}.
We see how the value of $\dm[41]$ shifts the peak, while the mixing angle adjusts its strength. 
%TODO: more? mention that Pmm=1 at TeV for 3+0
